---
title: "pyNMF-experiments-Dmel-carcass-data"
author: "snikumbh"
date: "`r Sys.time()`"
output:
  html_document: 
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    self_contained: no
    toc: yes
    code_folding: hide
    toc_float: true
editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
# setting echo=TRUE here leads to not saving any files in the corresponding folder  
knitr::opts_chunk$set(echo = TRUE, progress = TRUE)
processStart <- Sys.time()
```

```{r setup-two, echo=FALSE}
seed_val <- 11992288
set.seed(seed_val)

```

```{r setup-paths, echo=FALSE}
source('set_path_vars.R')
source('check_package_requirement.R')
# Path elsewhere
project_home_path <- '~/projects/clustering-promoter-architectures/promArch'
# Path on the server
# project_home_path <- '/mnt/biggley/home/sarvesh/projects/promoter-architectures-all/clustering-promoter-architectures/promArch'

paths_list <- set_path_vars(project_home_path)

# Check packages
# Solution from https://stackoverflow.com/a/23286257
list_of_packages <- c("tidyr",# 
                      "dplyr",#
                      "stringr",# gsub
                      "readr",# read_csv for reading csv files
                      "reticulate",# Sourcing python functions
                      "gplots",# heatmap.2 function
                      "RColorBrewer",# Custom color palette
                      "ggseqlogo",# SeqLogos with ggplot2
                      "ggplot2",# plotting 
                      "MASS",# ginv function for pseudoinverse
                      "BBmisc",# rowLapply
                      "cvTools",# cross-validation 
                      "tibble",
                      "purrr",#cross_df function
                      "pkgmaker", #source_files
                      "reshape2", #melt function
                      "heatmap3",
                      "ggplot2",
                      "cluster"
)
invisible(check_package_requirement(list_of_packages = list_of_packages))
```

```{r source-files, echo=FALSE}
source(file.path(paths_list$src_path, 'reqd_libraries.R'))
src_these <- paste0( file.path(paths_list$src_path, c("pyNMF_model_selection_functions.R", 
               "make_sinuc_PWMs.R", 
               "generate_seq_info_matrix.R", 
               "sparsify_mat.R",
               "plot_heatmap.R",
               "viz_all_factors_as_seqlogo.R",
               "viz_all_factors_as_heatmap.R",
               "viz_all_factors_in_combined_heatmaps_seqlogos.R",
               "plot_ggseqlogo.R",
               "plot_ggheatmap.R",
               "represent_matrix_of_acgt.R",
               "prepare_data_from_FASTA.R",
               "choose_clusters.R",
               "get_clusters.R",
               "plot_arch_for_clusters.R"
            )))
source_files(src_these)
source_python(file.path(paths_list$src_path, 'perform_nmf.py'))
```
# Dmel-carcass-data

## Fetch data

```{r sinuc-profiles-synthetic-dataset-2, echo=FALSE}

# Simply get filename and create the feature matrix
inputFastaFilename <- file.path(paths_list$drosophila_data_path, 'Dmel_carcass.fa')
tss.seqs_o <- prepare_data_from_FASTA(inputFastaFilename)
nSeqs <- ncol(tss.seqs_o)
positions <- setdiff(seq(-100,100), 0)
sinuc <- c('A', 'C', 'G', 'T')
tss.seqs <- tss.seqs_o
# Randomize the sequence order
#changed_order <- sample.int(nSeqs, nSeqs, replace = FALSE)
#tss.seqs <- tss.seqs_o[ , changed_order]

```

```{r plot-seqs, echo=FALSE}
# Represent sequences as heatmap

represent_matrix_of_acgt(tss.seqs, positions = positions)

```


## Run NMF and model selection

As is obvious, we need to fix the optimum number of factors before proceding to factorize. 
Additionally, as can be noted from the objective function of NMF (see equation above), some parameters determine the quality of factorization. 
For example, in the above, these are alpha and l1_ratio.

In order to determine the best performing model among the numerous possibilities, we perform a model selection procedure. 
We tune the parameters by 5-fold cross-validation in which we test for the reconstruction error of each model.
Note that the number of folds can be specified by the user.

In the following, we test the following range of values for $K$ and $\alpha$.

$K = \{3, 30\}$
  
$\alpha = 2^{-6,\ldots,3}$
  
```{r pyNMF-on-data, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
## Random ordered data
data_mat <- tss.seqs
##
param_ranges <- NULL
param_ranges$alphaBase <- 0
param_ranges$alphaPow <- 0
param_ranges$k_vals <- seq(3, 100)
kFolds <- 5
nCoresUse <- 8
seed_val <- 10208090
cat("Choosing optimal K\n")
# Fix number of patterns (K)
start <- Sys.time()
model_selectK <- invisible(cv_model_select_pyNMF(X=tss.seqs, 
                                                 param_ranges=param_ranges, 
                                                 kFolds=kFolds, 
                                                 parallel=FALSE, 
                                                 nCores=nCoresUse, 
                                                 seed=seed_val,
                                                 logfile=
                                                   file.path(paths_list$src_path,                                                 paste0('pyNMF-experiments-Dmel-carcass-data-worker-output-chooseK-', 
                                                                                                                                         paste(Sys.Date(), format(Sys.time(), "%X"), sep='-'),                                                           '.txt')
                                                   ),
                                                 set_verbose = 0
)
)
print(Sys.time() - start)

best_k <- get_best_k(model_selectK)
q2_means_by_k_vals <- get_q2_aggregates_chosen_var(model_selectK, model_selectK$k_vals, mean)
q2_threshold <- get_q2_threshold_by_k(model_selectK)
Q2vsK <- plot_cv_K(q2_means_by_k_vals)
print(Q2vsK)
#
param_ranges <- NULL
param_ranges$alphaBase <- 2
param_ranges$alphaPow <- seq(-6, 3)
param_ranges$k_vals <- best_k
seed_val <- 10208090
start <- Sys.time()
chooseAlpha_logfilename <- file.path(paths_list$src_path, paste0('pyNMF-experiments-Dmel-carcass-data-worker-output-chooseAlpha-',                                                               paste(Sys.Date(), format(Sys.time(), "%X"), sep='-'), '.txt')
)
model_selectAlpha <- invisible(cv_model_select_pyNMF(X=tss.seqs,
                                                     param_ranges=param_ranges,
                                                     kFolds=kFolds,
                                                     parallel=FALSE,
                                                     nCores=nCoresUse,
                                                     seed=seed_val,
                                                     logfile=chooseAlpha_logfilename,
                                                     set_verbose = 0
)
)
print(Sys.time() - start)

best_alpha <- get_best_alpha(for_alpha = model_selectAlpha, for_k = model_selectK, min_or_max = min)
q2_means_by_alpha <- get_q2_aggregates_chosen_var(model_selectAlpha, model_selectAlpha$alpha, mean)
#
cat(paste0("## Best K: ", best_k, ", Best Alpha: ", best_alpha, '\n\n'))
#
Q2vsAlpha <- plot_cv_Alpha(q2_means_by_alpha, q2_threshold)
print(Q2vsAlpha)
#
# 
```

```{r final-nmf-on-data, echo=FALSE}
cat("Performing NMF on the complete data...")
seed_val <- 10208090
start <- Sys.time()
result <- perform_nmf(tss.seqs, 
                      nPatterns = as.integer(best_k),
                      nIter = as.integer(1000),
                      givenAlpha = best_alpha,
                      givenL1_ratio = 1,
                      seed_val = as.integer(seed_val)
)
cat("done\n")
print(Sys.time() - start)

```



```{r fetch-factorized-matrices, echo=FALSE}
featuresMatrix <- result[[1]]
samplesMatrix <- result[[2]]

```


```{r sparsify-feature-matrix, echo=FALSE, message=FALSE, warning=FALSE}
cat(paste0("Number of factors: ", best_k, "\n"))
# Sparsify features
cat("Sparsifying features matrix...")
spFeaturesMatrix <- sparsify_mat(featuresMatrix)
cat("done\n")
#
```

<!-- Combined plots -- Sparsified Fatures Matrix -->
<!-- ```{r plot-patterns-as-PWMs-sparse, fig.height=5, fig.width=25, echo=FALSE, message=FALSE, warning=FALSE} -->
<!-- # Plot here and also save as PDF -- both done by the function -->
<!-- # filename_heatmaps <- file.path(paths_list$synthetic_data_results_path, "architectures-as-patterns-pyNMF-sparsified-heatmaps.pdf") -->
<!-- # viz_all_factors_as_heatmap(spFeaturesMatrix) -->
<!-- # # -->
<!-- # filename_seqlogos <- file.path(paths_list$synthetic_data_results_path, "architectures-as-patterns-pyNMF-sparsified.pdf") -->
<!-- # viz_all_factors_as_seqlogo(spFeaturesMatrix) -->
<!-- #  -->
<!-- #viz_all_factors_in_combined_heatmaps_seqlogos(spFeaturesMatrix) -->
<!-- ``` -->





# Combined plots -- Non-sparsified Features Matrix
```{r combined-plot-patterns-as-PWMs-non-sparse, fig.height=5, fig.width=25, echo=FALSE, warning=FALSE, message=FALSE}
# Plot here and also save as PDF -- both done in the function
# filename_heatmaps <- file.path(paths_list$synthetic_data_results_path, "architectures-as-patterns-pyNMF-sparsified-heatmaps.pdf")
# viz_all_factors_as_heatmap(featuresMatrix)
# #
# filename <- file.path(paths_list$synthetic_data_results_path, "architectures-as-patterns-pyNMF-non-sparsified.pdf")
# viz_all_factors_as_seqlogo(featuresMatrix)
#
viz_all_factors_in_combined_heatmaps_seqlogos(featuresMatrix, add_pseudo_counts = F)
```



```{r Cluster-sequences, echo=FALSE, warning=FALSE, message=FALSE}


final_nClust <- choose_clusters(t(samplesMatrix), clustMethod = "kmeans", nCluster_vals_test = seq(3,50))
cat("Number of clusters: ", final_nClust, "\n")

clustering_sol <- get_clusters(t(samplesMatrix), clustMethod = "kmeans", final_nClust)

```


```{r plot-seqs-post-nmf, fig.align = "center", echo=FALSE}
# Represent sequences as heatmap
# represent_matrix_of_acgt(tss.seqs, positions = positions, plot.title = "Given")
# represent_matrix_of_acgt(tss.seqs[ , ord_hm2$colInd], positions, plot.title = "After NMF")
# represent_matrix_of_acgt(tss.seqs, positions = positions, plot.title = "Given")
# # represent_matrix_of_acgt(tss.seqs[ , ss_hclust], positions = positions, plot.title = "hclust")
# # represent_matrix_of_acgt(tss.seqs[ , ord_hm2$colInd], positions, plot.title = "ord_hm2")
represent_matrix_of_acgt(tss.seqs[ , unlist(clustering_sol$reordering_idx)], positions, plot.title = "NMF+kmeans")


```

# Report architectures characteristic of clusters

```{r report-arch, fig.height=7, fig.width=25, echo=FALSE, warning=FALSE, message=FALSE}
# source(file.path(paths_list$src_path, "plot_arch_for_clusters.R"))

plot_arch_for_clusters(samplesMatrix, featuresMatrix, final_nClust, clustering_sol, tss.seqs, positions)

sprintf("Time taken from start to end: %s", format(Sys.time() - processStart, digits = 5, trim = F))
```

# Conclusion
NMF + K-means on the given data successfully identifies different clusters of sequences with characteristic architectures.




# Session Info
```{r session_info, include=TRUE, echo=TRUE, results='markup'}
sessionInfo()
```


